---
title: "Fixed Effects Model"
format: html
---

```{r}
#| child: "_chunk-opt.qmd"
#| include: false
```

Consider the following individual fixed effects model (at the level of the observation):

$$
\begin{aligned}
y_{it} &= \alpha_i + \bbeta'\bx_{it} + u_{it}, \quad t=1,2,\ldots,T, \\
\text{or}\quad y_{it} &= \alpha_i + \beta_1 x_{it1} + \beta_2 x_{it2} + \cdots + \beta_K x_{itK} + u_{it}.
\end{aligned}
$$  {#eq-within-lecture-random-intercept}
There are $K$ regressors in $\bx_{it},$ **NOT** including an intercept.

We assume strict exogeneity:

$$
\E[\bu_i|\bX_i]=0.
$$

Throughout, we will assume that the number $N$ of cross-sectional units is potentially large, while the number $T$ of data points per unit is fixed. This setup is more reflective of typical micropanels. 

**Alternative notations**:

At the level of the individual, the model is written as:

$$
\by_i = \bX_i \bbeta + \bold{1} \alpha_i + \bvarepsilon_i  ,
$$
where $\bold{1}$ is a $T\times 1$ column of ones.

Equation for the full sample:

$$
\begin{bmatrix}
\by_{1}  \\
\by_{2}  \\
\vdots \\
\by_{N}  
\end{bmatrix}  = 
\begin{bmatrix}
\bX_{1}  \\
\bX_{2}  \\
\vdots \\
\bX_{N}  
\end{bmatrix} \bbeta + \begin{bmatrix}
\bold{1} & 0  & \cdots & 0 \\
0 & \bold{1} & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0  & \cdots & \bold{1} \\
\end{bmatrix} 
\begin{bmatrix}
\alpha_{1}  \\
\alpha_{2}  \\
\vdots \\
\alpha_{N} 
\end{bmatrix} 
+ \begin{bmatrix}
\bvarepsilon_{1}  \\
\bvarepsilon_{2}  \\
\vdots \\
\bvarepsilon_{M}  
\end{bmatrix}
$$
or 

$$
\begin{split}
\by &= \begin{bmatrix} \bX & \bd_1 & \bd_2 & \cdots & \bd_n \end{bmatrix} 
\begin{bmatrix}
\bbeta \\
\balpha
\end{bmatrix} + \bvarepsilon \\
&= \bX\bbeta + \bD\balpha + \bvarepsilon  ,
\end{split}
$$
where $\bd_i$ is a dummy variable indicating the $i$-th unit, and 

$$
\underset{(NT\times N)}{\bD} = \begin{bmatrix}\bd_1 & \bd_2 & \cdots & \bd_n \end{bmatrix} .
$$

## Within Estimator

Within (or "fixed effects") estimation of the random intercept model has two steps:

1. Apply the within transformation to each unit.
2. Apply OLS to the resulting pooled data.

### Step 1: Within Transformation

To perform the within transformation, we first average the equations for unit $i$ across $t$. Label the average of $y_{it}$ across $t$ for unit $i$ as  

$$
y_{i\cdot} = \frac{1}{T} \sum_{t=1}^{T} y_{it}.
$$

The averaged outcome $y_{i\cdot}$ satisfies the averaged (or "cross-sectional") equation

$$
y_{i\cdot} = \alpha_i + \bbeta' \bx_{i\cdot} + u_{i\cdot},
$$ {#eq-within-lecture-avg}
where $\bx_{i\cdot}$ and $u_{i\cdot}$ are defined analogously to $y_{i\cdot}$.

Define the within-transformed equation by subtracting this averaged equation (-@eq-within-lecture-avg) from the original equation (-@eq-within-lecture-random-intercept) for each $t$:

$$
y_{it} - y_{i\cdot} = (\alpha_i - \alpha_i) + \bbeta'(\bx_{it} - \bx_{i\cdot}) + (u_{it} - u_{i\cdot}) 
$$
or 
$$
\tilde{y}_{it} = \bbeta' \tilde{\bx}_{it} + \tilde{u}_{it}.
$${#eq-within_transformed}

where $\tilde{y}_{it} = y_{it} - y_{i\cdot}$ is the deviations from the group means, and similarly for $\tilde{\bx}_{it}$ and $\tilde{u}_{it}.$

Under @eq-within-lecture-random-intercept, <span style='color:#008B45'>the within transformation eliminates the individual random intercepts $\alpha_i$</span>. All average differences in $y$'s or $\bx$'s between individuals have been wiped out.
@eq-within_transformed now looks like a regular homogeneous regression. 

### Step 2: OLS on the Within-Transformed Equation

The within (fixed effects) estimator is obtained by simply pooling the data across $i$ and $t$ in @eq-within_transformed and applying OLS to it. Specifically, the estimator is given by

$$
\hat{\bbeta}^W = \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \sum_{i=1}^{N} \tilde{\bX}_i \tilde{\mathbf{y}}_i.
$${#eq-within_representation_random_intercept}


### Properties

The within estimator enjoys several desirable properties if the random intercept model reflects the underlying causal model. Most of these properties can be derived from its sampling error representation
$$
\hat{\bbeta}^W = \bbeta + \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \sum_{i=1}^{N} \tilde{\bX}_i \tilde{\bu}_i.
$${#eq-within-sampling-error-homogeneous}

- $\hat{\bbeta}^W$ is unbiased for $\bbeta$. To show this, it is sufficient to notice that strict exogeneity of $\bu_i$ with respect to $\bX_i$ implies strict exogeneity of $\tilde{\bu}_i$ with respect to $\tilde{\bX}_i$:
$$
\E[\tilde{\bu}_i|\tilde{\bX}_i] = 
\E[\E[\tilde{\bu}_i|\bX_i]|\tilde{\bX}_i] = 0.
$$
It follows that the mean of the second term in @eq-within-sampling-error-homogeneous is 0, and so
 $$
\E[\hat{\bbeta}^W] = \bbeta.
 $$

- $\hat{\bbeta}^W$ is consistent for $\bbeta$ and asymptotically normal, provided a standard rank condition holds for $\tilde{\bX}_i$:
$$
\hat{\bbeta}^W \xrightarrow{p} \bbeta, \quad \sqrt{N}(\hat{\bbeta}^W - \bbeta) \Rightarrow N(0, \Sigma).  
$${#eq-within_asymptotic}


Since $\bbeta$ is the average coefficient vector in this homogeneous model, we conclude that the within estimator consistently estimates average coefficients under the random intercept model ([-@eq-within-lecture-random-intercept]).

## Between-groups Estimator

The Between-groups estimator (or between estimator) is obtained by regressing the time averaged variables on each other (-@eq-within-lecture-avg) (where we include an intercept, $\beta_0$) using OLS regression. 
$$
y_{i\cdot} = \alpha_i + \bbeta' \bx_{i\cdot} + u_{i\cdot},
$$
This is a cross-sectional regression, with a sample size being the number of entities.

Note that the between estimator is **biased** when $\alpha_i$ is correlated with $\bx_{i\cdot}.$ If we think $\alpha_i$ is uncorrelated with $\bx_{i\cdot},$ it is better to use the random effects estimator.

It can be suitable for research questions that specifically address variation between different entities rather than changes within entities over time. One caveat is that the Between-groups estimator *overlooks important information about how variables change over time*.

To obtain unbiased estimates, it is crucial to ensure the assumption of zero correlation between the error term and averaged independent variables.

```{r}
library(plm)
library(AER) 
data(Grunfeld) 

# Estimate Between model
model_between <- plm(invest ~ value + capital, 
                      data = Grunfeld,
                      index = c("firm", "year"),
                      model = "between")
# Estimate Within model
model_within <- plm(invest ~ value + capital, 
                      data = Grunfeld,
                      index = c("firm", "year"),
                      model = "within")
```

```{r results='asis'}
library(stargazer)
stargazer(model_between, model_within, 
          column.labels = c("Between", "Within"),
          type="html", digits=2, 
          notes = "<span>&#42;</span>: p<0.1; <span>&#42;&#42;</span>: <strong>p<0.05</strong>; <span>&#42;&#42;&#42;</span>: p<0.01 <br> Standard errors in parentheses.",
          notes.append = F)
```




## References {.unlisted .unnumbered}

- Vladislav Morozov, Econometrics with Unobserved Heterogeneity, Course material, <https://vladislav-morozov.github.io/econometrics-heterogeneity/linear/linear-within-estimator.html#recall-within-estimator-in-the-random-intercept-model>
- Vladislav Morozov, GitHub course repository, <https://github.com/vladislav-morozov/econometrics-heterogeneity/blob/main/src/linear/linear-within-estimator.qmd>






