---
title: "Fixed Effects Model"
format: html
---

```{r}
#| child: "_chunk-opt.qmd"
#| include: false
```

Consider the following individual fixed effects model (at the level of the observation):

$$
\begin{aligned}
y_{it} &= \alpha_i + \bbeta'\bx_{it} + u_{it}, \quad t=1,2,\ldots,T, \\
\text{or}\quad y_{it} &= \alpha_i + \beta_1 x_{it1} + \beta_2 x_{it2} + \cdots + \beta_K x_{itK} + u_{it}.
\end{aligned}
$$  {#eq-within-lecture-random-intercept}
There are $K$ regressors in $\bx_{it},$ **NOT** including an intercept.

We assume strict exogeneity:

$$
\E[\bu_i|\bX_i]=0.
$$

Throughout, we will assume that the number $N$ of cross-sectional units is potentially large, while the number $T$ of data points per unit is fixed. This setup is more reflective of typical micropanels. 

**Alternative notations**:

Stacking the $T$ equations at the level of the individual, the model is written as:

$$
\by_i = \bX_i \bbeta + \bold{1} \alpha_i + \bu_i  ,
$$
where $\bold{1}$ is a $T\times 1$ column of ones.

Equation for the full sample:

$$
\begin{bmatrix}
\by_{1}  \\
\by_{2}  \\
\vdots \\
\by_{N}  
\end{bmatrix}  = 
\begin{bmatrix}
\bX_{1}  \\
\bX_{2}  \\
\vdots \\
\bX_{N}  
\end{bmatrix} \bbeta + \begin{bmatrix}
\bold{1} & 0  & \cdots & 0 \\
0 & \bold{1} & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0  & \cdots & \bold{1} \\
\end{bmatrix} 
\begin{bmatrix}
\alpha_{1}  \\
\alpha_{2}  \\
\vdots \\
\alpha_{N} 
\end{bmatrix} 
+ \begin{bmatrix}
\bu_{1}  \\
\bu_{2}  \\
\vdots \\
\bu_{N}  
\end{bmatrix}
$$
or 

$$
\begin{split}
\by &= \begin{bmatrix} \bX & \bd_1 & \bd_2 & \cdots & \bd_n \end{bmatrix} 
\begin{bmatrix}
\bbeta \\
\balpha
\end{bmatrix} + \bu \\
&= \bX\bbeta + \bD\balpha + \bu  ,
\end{split}
$$
where $\bd_i$ is a dummy variable indicating the $i$-th unit, and 

$$
\underset{(NT\times N)}{\bD} = \begin{bmatrix}\bd_1 & \bd_2 & \cdots & \bd_n \end{bmatrix} .
$$

## Within Estimator

Within (or "fixed effects") estimation of the random intercept model has two steps:

1. Apply the within transformation to each unit.
2. Apply OLS to the resulting pooled data.

### Step 1: Within Transformation

To perform the within transformation, we first average the equations for unit $i$ across $t$. Label the average of $y_{it}$ across $t$ for unit $i$ as  

$$
\overline{y}_{i} = \frac{1}{T} \sum_{t=1}^{T} y_{it}.
$$

The averaged outcome $\overline{y}_{i}$ satisfies the averaged (or "cross-sectional") equation

$$
\overline{y}_{i} = \alpha_i + \bbeta' \overline{\bx}_{i} + \overline{u}_{i},
$$ {#eq-within-lecture-avg}
where $\overline{\bx}_{i}$ and $\overline{u}_{i}$ are defined analogously to $\overline{y}_{i}$.

Define the within-transformed equation by subtracting this averaged equation (-@eq-within-lecture-avg) from the original equation (-@eq-within-lecture-random-intercept) for each $t$:

$$
y_{it} - \overline{y}_{i} = (\alpha_i - \alpha_i) + \bbeta'(\bx_{it} - \overline{\bx}_{i}) + (u_{it} - \overline{u}_{i}) 
$$
or 
$$
\widetilde{y}_{it} = \bbeta' \widetilde{\bx}_{it} + \widetilde{u}_{it}.
$${#eq-within_transformed}

where $\widetilde{y}_{it} = y_{it} - \overline{y}_{i}$ is the deviations from the group means, and similarly for $\widetilde{\bx}_{it}$ and $\widetilde{u}_{it}.$

Under @eq-within-lecture-random-intercept, <span style='color:#008B45'>the within transformation eliminates the individual random intercepts $\alpha_i$</span>. All average differences in $y$'s or $\bx$'s between individuals have been wiped out.
@eq-within_transformed now looks like a regular homogeneous regression. 

### Step 2: OLS on the Within-Transformed Equation

The within (fixed effects) estimator is obtained by simply pooling the data across $i$ and $t$ in @eq-within_transformed and applying OLS to it. Specifically, the estimator is given by

$$
\begin{aligned}
\hat{\bbeta}^W &= \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \sum_{i=1}^{N}  \left( \tilde{\bX}_i'\, \tilde{\mathbf{y}}_i \right) \\
&= \left(\sum_{i=1}^{N} \sum_{=1}^{T} \tilde{\bx}_{it} \tilde{\bx}_{it}'  \right)^{-1} \left(\sum_{i=1}^{N}\sum_{=1}^{T} \tilde{\bx}_{it}\, \tilde{\mathbf{y}}_{it} \right) .
\end{aligned}
$$ {#eq-within_representation_random_intercept}

It is called the within estimator because it uses the time variation within each individual.

### Properties

The within estimator enjoys several desirable properties if the random intercept model reflects the underlying causal model. Most of these properties can be derived from its sampling error representation
$$
\hat{\bbeta}^W = \bbeta + \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bu}_i.
$${#eq-within-sampling-error-homogeneous}

- $\hat{\bbeta}^W$ is unbiased for $\bbeta$. To show this, it is sufficient to notice that strict exogeneity of $\bu_i$ with respect to $\bX_i$ implies strict exogeneity of $\tilde{\bu}_i$ with respect to $\tilde{\bX}_i$:
$$
\E[\tilde{\bu}_i|\tilde{\bX}_i] = 
\E[\E[\tilde{\bu}_i|\bX_i]|\tilde{\bX}_i] = 0.
$$
It follows that the mean of the second term in @eq-within-sampling-error-homogeneous is 0, and so
 $$
\E[\hat{\bbeta}^W] = \bbeta.
 $$


- The variance of $\hat{\bbeta}^W$ is

    $$
    \bV_{\hat{\bbeta}^W} = \var(\hat{\bbeta}^W\mid \bX) = \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \left( \sum_{i=1}^{N} \tilde{\bX}_i'\, \E[\tilde{\bu}_i \tilde{\bu}_i'\mid \bX_i] \,\tilde{\bX}_i \right) \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} .
    $$
    Let 
    $$
    \bSigma_i = \E[\tilde{\bu}_i \tilde{\bu}_i'\mid \bX_i] 
    $$
    be the $T\times T$ conditional matrix of the idiosyncratic errors. We can rewrite $\bV_{\hat{\bbeta}^W}$ as
    
    $$
    \bV_{\hat{\bbeta}^W} = \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \left( \sum_{i=1}^{N} \tilde{\bX}_i'\, \bSigma_i \,\tilde{\bX}_i \right) \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} .
    $$
    
    If we assume the idiosyncratic errors are homoskedastic and serially uncorrelated:
    
    $$
    \begin{aligned}
    \E(u_{it}^2 \mid \bX_i) &= \sigma^2_u \\
    \E(u_{is}u_{it} \mid \bX_i) &= 0 \quad (x\ne t)
    \end{aligned}
    $$
    
    In this case, 
    
    $$
    \bSigma_i = \mathbf I_i \sigma^2_u .
    $$
    
    $\bV_{\hat{\bbeta}^W}$ simplifies to
    
    $$
    \bV_{\hat{\bbeta}^W}^0 = \sigma^2_u \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} .
    $$

- $\hat{\bbeta}^W$ is consistent for $\bbeta$ and asymptotically normal, provided a standard rank condition holds for $\tilde{\bX}_i$:
$$
\hat{\bbeta}^W \xrightarrow{p} \bbeta, \quad \sqrt{N}(\hat{\bbeta}^W - \bbeta) \Rightarrow N(0, \bV_{\hat{\bbeta}^W}).  
$${#eq-within_asymptotic}


Since $\bbeta$ is the average coefficient vector in this homogeneous model, we conclude that the within estimator consistently estimates average coefficients under the random intercept model ([-@eq-within-lecture-random-intercept]).


- Covariance matrix estimation

    An estimator for $\bV_{\hat{\bbeta}^W}^0$ is
    
    $$
    \widehat{\bV}_{\hat{\bbeta}^W}^0 = \hat{\sigma}^2_u \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1}
    $$
    where $\hat{\sigma}^2_u$ is an unbiased estimator of $\sigma^2_u$ given by:
    
    $$
    \hat{\sigma}^2_u = \frac{\sum_{i=1}^{N}\sum_{t=1}^{T}\widehat{\widetilde{u}}_{it}^2}{NT-N-K} .
    $$


- A covariance matrix estimator which allows $u_{it}$ to be heteroskedastic and serially correlated across $t$ is the cluster-robust covariance matrix estimator, clustered by individual

$$
\widehat{\bV}_{\hat{\bbeta}^W}^{\text{cluster}} = \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} \left( \sum_{i=1}^{N} \tilde{\bX}_i'\, \widehat{\widetilde{\bu}}_{i}\widehat{\widetilde{\bu}}_{i}' \,\tilde{\bX}_i \right) \left(\sum_{i=1}^{N} \tilde{\bX}_i' \tilde{\bX}_i \right)^{-1} 
$$

## Between-groups Estimator

The Between-groups estimator (or between estimator) is obtained by regressing the **time averaged** variables on each other (-@eq-within-lecture-avg) (where we include an intercept, $\beta_0$) using OLS regression. 
$$
\overline{y}_{i} =\bbeta' \overline{\bx}_{i} + (\alpha_i + \overline{u}_{i}) , \quad i=1,\ldots,N.
$$
This is a cross-sectional regression, with a sample size being the number of entities.

The Between Groups estimator is given by

$$
\hat{\bbeta}^{\text{B}} = \left( \sum_{i=1}^N \overline{\bx}_{i}\overline{\bx}_{i}'\right)^{-1} \left( \sum_{i=1}^N \overline{\bx}_{i}\overline{y}_{i}\right)
$$

Note that the between estimator is **biased** when $\alpha_i$ is correlated with $\bx_{i\cdot}$ ($\bx_{i\cdot}$ endogenous). 

It can be suitable for research questions that specifically address variation between different entities rather than changes within entities over time. One caveat is that the Between-groups estimator *overlooks important information about how variables change over time*.

To obtain unbiased estimates, it is crucial to ensure the assumption of zero correlation between the error term and averaged independent variables.

$$
\E[\bx_{i\cdot} \alpha_i] = 0 .
$$
Between Groups estimator is not efficient. In practice, it is only used to obtain an estimate of $\sigma_{\alpha}^2$ when implementing feasible GLS.

If we think $\alpha_i$ is uncorrelated with $\bx_{i\cdot},$ it is better to use the random effects estimator.

```{r}
library(plm)
library(AER) 
data(Grunfeld) 

# Estimate Between model
model_between <- plm(invest ~ value + capital, 
                      data = Grunfeld,
                      index = c("firm", "year"),
                      model = "between")
# Estimate Within model
model_within <- plm(invest ~ value + capital, 
                      data = Grunfeld,
                      index = c("firm", "year"),
                      model = "within")
```

```{r results='asis'}
library(stargazer)
stargazer(model_between, model_within, 
          column.labels = c("Between", "Within"),
          type="html", digits=2, 
          notes = "<span>&#42;</span>: p<0.1; <span>&#42;&#42;</span>: <strong>p<0.05</strong>; <span>&#42;&#42;&#42;</span>: p<0.01 <br> Standard errors in parentheses.",
          notes.append = F)
```




## References {.unlisted .unnumbered}

- Vladislav Morozov, Econometrics with Unobserved Heterogeneity, Course material, <https://vladislav-morozov.github.io/econometrics-heterogeneity/linear/linear-within-estimator.html#recall-within-estimator-in-the-random-intercept-model>
- Vladislav Morozov, GitHub course repository, <https://github.com/vladislav-morozov/econometrics-heterogeneity/blob/main/src/linear/linear-within-estimator.qmd>






