---
title: "TS Examples"
format: html
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: load-packages
#| include: false
library(knitr) # load packages
library(broom)
library(kableExtra)
library(tidyverse)
library(latex2exp)
library(stargazer)
library(xts)
library(wooldridge)
```

```{r}
#| child: "_chunk-opt.qmd"
#| include: false
```


Textbook: Chapter 11, **Introductory Econometrics: A Modern Approach**, 7e by Jeffrey M. Wooldridge

Summary notes by Marius v. Oordt: <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3401712>

## `Example 11.6:` Fertility and Personal Exemption

`gfr`: general fertility rate

`pe`: personal exemption

```{r}
data('fertil3')
fertility_diff <- lm(diff(gfr) ~ diff(pe), data = fertil3)
fertility_lag <- lm(diff(gfr) ~ diff(pe) + diff(pe_1) + diff(pe_2), data = fertil3)
```

```{r results='asis', echo=FALSE}
stargazer(type = "html", fertility_diff, fertility_lag, single.row = TRUE, header = FALSE, digits=5, notes = "<span>&#42;</span>: p<0.1; <span>&#42;&#42;</span>: <strong>p<0.05</strong>; <span>&#42;&#42;&#42;</span>: p<0.01 <br> Standard errors in parentheses.", notes.append = FALSE)
```

The first regression uses first differences:

$$
\begin{aligned}
\Delta\widehat{gfr} &= -.785 - .043\, \Delta pe \\
&\phantom{=}\;\; (.502)\;\; (.028) \\
n &= 71, R^2=.032, \bar{R^2} = .018.
\end{aligned}
$$  {#eq-gfr_static}

The estimates indicate an increase in $pe$ lowers $gfr$ contemporaneously, although the estimate is not statistically different from zero at the 5% level.

If we add two lags of $\Delta pe,$ things improve:

$$
\begin{aligned}
\Delta\widehat{gfr} &= -.964 - .036\, \Delta pe - .014\, \Delta pe_{-1} + .110\, \Delta pe_{-2} \\
&\phantom{=}\;\; (.468)\quad (.027) \qquad\; (.028) \qquad\quad\; (.027)\\
n &= 69, R^2=.232, \bar{R^2} = .197.
\end{aligned}
$$ {#eq-gfr_lag}

We call model (-@eq-gfr_lag) an **finite distributed lag** (FDL) model of order two. A more general specification is

$$
y_t = \alpha_0 + \delta_0 z_t + \delta_1 z_{t-1} + \delta_2 z_{t-2} + u_t .
$$

Even though $\Delta pe$ and $\Delta pe_{-1}$ have negative coefficients, their coefficients are small and jointly insignificant ($p\text{-value}=.28,$ see Anova test below). 

```{r}
# Compare the restricted with the full model
fertility_lag2 <- lm(diff(gfr) ~ diff(pe_2), data = fertil3)
anova(fertility_lag2, fertility_lag)
```

The second lag ($\Delta pe_{-2}$) is very significant and indicates a positive relationship between changes in $pe$ and subsequent changes in $gfr$ two years hence. This makes more sense than having a contemporaneous effect.


### `Example 11.8`

In this example, we want to test whether the **Finite Distributed Lag model** (-@eq-gfr_lag) for $\Delta\widehat{gfr}$ and $\Delta pe$ is **dynamically complete**.

Being dynamically complete indicates that neither lags of $\Delta\widehat{gfr}$ nor further lags of $\Delta pe$ should appear in the equation. Mathematically, given the following finite distributed lag model:

$$
\Delta gfr_t = \beta_0 + \beta_1\Delta pe_t + \beta_2\Delta pe_{t-1} + \beta_3 \Delta pe_{t-2} + u_t .
$$
Rewrite it as

$$
\begin{aligned}
\Delta gfr_t &= \beta_0 + \beta_1x_{t1} + \beta_2x_{t2} + \beta_3 x_{t3} + u_t \\
y_t &= \bx_t'\bbeta + u_t
\end{aligned}
$$
where the explanatory variables $\bx_t=(x_{t1}, x_{t2}, x_{t3})' = (\Delta pe_t, \Delta pe_{t-2}, \Delta pe_{t-3})'$ and the dependent variable $y_t=\Delta gfr_t.$

A dynamically complete model requires the following condition:

$$
\E(u_t\mid \bx_t, y_{t-1}, \bx_{t-1}, \ldots) = 0.
$$ {#eq-dc1}
Written in terms of $y_t,$

$$
\E(y_t\mid \bx_t, y_{t-1}, \bx_{t-1}, \ldots) = \E(y_t\mid \bx_t).
$$  {#eq-dc2}

We can test for dynamic completeness by adding $\Delta gfr_{t-1}.$

```{r}
fertility_lag_dep <- lm(diff(gfr) ~ lag(diff(gfr)) + diff(pe) + diff(pe_1) + diff(pe_2), data = fertil3)
tidy(fertility_lag_dep) %>% 
    knitr::kable(digits = 3) 
```

The coefficient estimate is .300 and its $t$ statistic is 2.84. Thus, the model is NOT dynamically complete in the sense of (-@eq-dc2).


The fact that (-@eq-gfr_lag) is not dynamically complete suggests that there may be serial correlation in the errors. We will need to test and correct for this.



## `Example 11.7:` Wages and Productivity


$$\log(hrwage_t) = \beta_0 + \beta_1\log(outphr_t) + \beta_2t + u_t$$
Data from the _Economic Report of the President, 1989_, Table B-47. The data are for the non-farm business sector.

```{r}
data("earns")
wage_time <- lm(lhrwage ~ loutphr + t, data = earns)
wage_diff <- lm(diff(lhrwage) ~ diff(loutphr), data = earns)
```

```{r results='asis', echo=FALSE}
stargazer(type = "html", wage_time, wage_diff,  single.row = TRUE, header = FALSE, digits=5, notes = "<span>&#42;</span>: p<0.1; <span>&#42;&#42;</span>: <strong>p<0.05</strong>; <span>&#42;&#42;&#42;</span>: p<0.01 <br> Standard errors in parentheses.", notes.append = FALSE)
```