---
title: "Phillips Curve"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| include: false
library(knitr) # load packages
library(broom)
library(kableExtra)
library(tidyverse)
library(latex2exp)
library(stargazer)
library(xts)
```

```{r}
#| child: "_chunk-opt.qmd"
#| include: false
```

## Static Phillips Curve

A static Phillips curve is given by:

$$
inf_t = \beta_0 + \beta_1\, unem_t + u_t,
$$

where $inf_t$ is the annual inflation rate and $unem_t$ is the unemployment rate.

This form of the Phillips curve assumes a constant natural rate of unemployment and constant inflationary expectations, and it can be used to study the contemporaneous tradeoff between inflation and unemployment.

```{r}
# data preview
data <- read.table("https://raw.githubusercontent.com/my1396/course_dataset/refs/heads/main/TableF5-2.txt", header = TRUE)
data <- data %>% 
    mutate(delta_infl = infl-lag(infl))
data %>% 
    head() %>% 
    knitr::kable(digits = 5) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, latex_options="scale_down") %>% 
    scroll_box(width = "100%")
```


```{r}
lm_phillips_stat <- lm(infl ~ unemp, data=data)
summary(lm_phillips_stat)
```

The simple regression estimates are

$$
\begin{aligned}
\widehat{inf}_t &= 2.22 + 0.30\, unem_t \\
&\phantom{={ }} (0.89)\;\; (.15)
\end{aligned}
$$
The regression indicates a positive relationship ($\hat{\beta}_1>0$) between inflation and unemployment at 5% significance level. 


There are some problems with this analysis that we cannot address in detail now. The Classical Linear Model assumptions do not hold. In addition, the static Phillips curve is probably not the best model for determining whether there is a short run tradeoff between inflation and unemployment. Macroeconomists generally prefer the **expectations augmented Phillips curve**, while we will see shortly.



## Expectations Augmented Phillips Curve

A linear version of the expectations augmented Phillips curve can be written as

$$
inf_t - inf^e_t = \beta_1 (unem_t - \mu_0) + e_t,
$$

where $\mu_0$ is the *natural rate of unemployment* and $inf^3_t$ is the *expected* rate of inflation formed in year $t-1.$

The difference between actual unemployment and the natural rate is called *cyclical unemployment*, while the difference between actual and expected
inflation is called *unanticipated inflation*.

The error term, $e_t$, is called a *supply shock* by macroeconomists.

To complete this model, we need to make an assumption about inflationary expectations.
Under adaptive expectations, the expected value of current inflation depends on recently observed inflation. A particularly simple formulation is that expected inflation this year is last year's inflation:

$$
inf^e_t = inf_{t-1}
$$

Under this assumption, we can write

$$
inf_t - inf_{t-1} = \beta_0 + \beta_1 \,unem_t  + e_t,
$$
or 

$$
\Delta  inf_t = \beta_0 + \beta_1 \,unem_t  + e_t,
$$
where $\Delta  inf_t = inf_t - inf_{t-1}$ and $\beta_0=-\beta_1\mu_0.$

```{r}
lm_phillips_aug <- lm(delta_infl ~ unemp, data)
summary(lm_phillips_aug)
```

The OLS estimates are

$$
\begin{aligned}
\Delta\widehat{inf}_t &= 0.52 - 0.09 \,unem_t \\
&\phantom{={}} (0.74) \;\; (0.13)
\end{aligned}
$$

$\hat{\beta}_1<0$ indicates a tradeoff between cyclical unemployment and unanticipated inflation. But the effect is statistically insignificant.

The natural employment rate is 5.78 percent.

$$
\mu_0 = -\frac{\beta_0}{\beta_1} = \frac{0.52}{0.09} \approx 5.78
$$



