---
title: "AR -- Example"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| include: false
library(knitr) # load packages
library(broom)
library(kableExtra)
library(tidyverse)
library(latex2exp)
library(stargazer)
library(xts)
```

```{r}
#| child: "_chunk-opt.qmd"
#| include: false
```

This script provides an example of autocorrelated residuals using Static Phillips Curve.

## Dataset Description

US Macroeconomics Data Set, Quarterly, 1950I to 2000IV, 204 Quarterly Observations \
Source: Department of Commerce, BEA website and [www.economagic.com](www.economagic.com)


| Feild Name | Definition                                                   |
| :--------- | :----------------------------------------------------------- |
| year       | Year                                                         |
| qtr        | Quarter                                                      |
| realgdp    | Real GDP (\$bil)                                             |
| realcons   | Real consumption expenditures                                |
| realinvs   | Real investment by private sector                            |
| realgovt   | Real government expenditures                                 |
| realdpi    | Real disposable personal income                              |
| cpi_u      | Consumer price index                                         |
| M1         | Nominal money stock                                          |
| tbilrate   | Quarterly average of month end 90 day t bill rate            |
| unemp      | Unemployment rate                                            |
| pop        | Population, mil. interpolate of year end figures using constant growth rate per quarter |
| infl       | Rate of inflation (First observation is missing)             |
| realint    | Ex post real interest rate = Tbilrate - Infl. (First observation is missing) |

```{r}
# data preview
data <- read.table("https://raw.githubusercontent.com/my1396/course_dataset/refs/heads/main/TableF5-2.txt", header = TRUE)
data <- data %>% 
    mutate(delta_infl = infl-lag(infl))
data %>% 
    head() %>% 
    knitr::kable(digits = 5) %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, latex_options="scale_down") %>% 
    scroll_box(width = "100%")
```


## Empirical Model

$$
\Delta I_t =  \beta_1 + \beta_2 u_t + \varepsilon_t
$$
where 

- $I_t$ is the inflation rate; $\Delta I_t = I_t - I_{t-1}$ is the first difference of the inflation rate;
- $u_t$ is the unemployment rate;
- $\varepsilon_t$ is the error term.

We remove the first two quarters due to missing value in the first observation and the change in the rate of inflation.


Regression result for OLS.

```{r}
#| results: asis
lm_phillips <- lm(delta_infl ~ unemp, data = data %>% tail(-2))
stargazer(lm_phillips, 
          type = "html", 
          title = "Phillips Curve Regression",
          notes = "<span>&#42;</span>: p<0.1; <span>&#42;&#42;</span>: <strong>p<0.05</strong>; <span>&#42;&#42;&#42;</span>: p<0.01 <br> Standard errors in parentheses.",
          notes.append = F)
```

```{r}
# variance-covariance matrix
vcov(lm_phillips)
```

**Autocorrelated residuals**

Plot the residuals.
```{r}
#| label: fig-ar-res
#| fig.cap: "Phillips Curve Deviations from Expected Inflation"
plot(lm_phillips$residuals, type="l")
```

@fig-ar-res shows striking negative autocorrelation.

Now we test the serial correlation of the residuals by regressing $\varepsilon_t$ on $\varepsilon_{t-1}$.
$$
\varepsilon_t = \phi\varepsilon_{t-1} + e_t
$$

```{r}
res <- tibble(
    res_t = lm_phillips$residuals,
    res_t1 = lag(lm_phillips$residuals))
lm_res <- lm(res_t ~ res_t1, data = res)
summary(lm_res)
```

The regression of the least squares residuals on their past values gives a slope of `r sprintf("%.4f", lm_res$coefficients[2])` with a highly significant $t$ ratio of `r sprintf("%.4f", summary(lm_res)$coef[2,3])`.
We thus conclude that the residuals in this models are highly negatively autocorrelated. 


Testing of serial correlation:

- Durbin-Watson (DW) test for AR(1)
- Breusch-Godfrey test for AR(q)

```{r}
library(lmtest)
dwtest(lm_phillips, alternative = "two.sided") # Durbin Watson test 
bgtest(lm_phillips, order=1) # Breusch-Godfrey test 
```

Both tests show strong evidence of AR(1) serial correlation in the errors.

One **consequence** of the serial correlated errors is that the standard error and $t$ statistics are not valid anymore. In the case if serial correlation, you can either 

- Transform the model to remove the serial correlation, or alternatively,
    
    FGLS (Feasible Genralized Least Squares), transform the original equation using, e.g., Cochrane-Orcutt or Prais-Winsten transformation.
    
- Use serial correlation-robust standard errors

    Heteroskedasticity and autocorrelation consistent, or HAC, standard errors.



--------------------------------------------------------------------------------

## References {.unlisted .unnumbered}

- Ex. 12.3, Chap 12 Serial Correlation, *Econometric Analysis*, Greene 5th Edition, pp 251.











